Do we need detailed demographic data to forecast plant population responses to climate change?

### 2014-12-29 manuscript/manuscript.Rmd
added climate tracking script
Added a climate tracking script, a la Jim Clark. Probably won’t use it
for this project, but could be interesting to look at all species at
all sites and examine climate tracking…
updated priors
new analysis update

@@

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76). The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-m2 quadrats using a pantograph. The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment). In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions. The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program. Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.

@@

\begin{align}
\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
\end{align}

@@

\begin{align}
\boldsymbol{\gamma, \beta} &\sim \text{Normal}(0, 1e-6) \\
\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
\boldsymbol{\sigma_{\phi}} &\sim e(\text{Gamma}(2, 0.5)) \\
\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}} &\sim \text{Gamma}(0.001, 0.001)
\end{align}

@@

### Population models
With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.

Results
-------
We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only. When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].

@@

gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
gIC <- gIC[133:148,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
gIC$X <- rep(coefNames, each=4)
gIC$rate <- "growth"
sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
sIC <- sIC[133:148,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
sIC$X <- rep(coefNames, each=4)
sIC$rate <- "survival"
rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
rIC <- rIC[81:96,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
rIC$X <- rep(coefNames, each=4)
rIC$rate <- "colonization"
plotIC <- rbind(gIC,sIC,rIC)
gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
  geom_hline(aes(yintercept=0), linetype="dashed")+
  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
  ylab("Standardized Coefficient Value")+
  xlab("Species")+
  facet_grid(X~.)+
  coord_flip()+
  theme_bw()+
  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
                     labels=c("recruitment", "growth", "survival"))+
#   guides(color=FALSE)+
  guides(color=guide_legend(title=""))+
  theme(legend.position="top")+
  ggtitle("Individual level")

@@

>>>>

*Keywords*: forecasting, climate change, grassland, integral projection model

@@

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment). In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions. The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program. Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.

@@

\begin{align}
\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k})
\end{align}

where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th climate covariate.

@@

\begin{align}
\boldsymbol{\gamma, \beta, \bar{\theta}} &\sim \text{Normal}(0, 1e^{-6}) \\
\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
\boldsymbol{\sigma_{\phi}} &\sim e^{(\text{Gamma}(2, 0.5))} \\
\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}, \tau, \zeta} &\sim \text{Gamma}(0.001, 0.001)
\end{align}

@@

We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only. When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].

### Population models
With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models. We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data. Both models take the general form:

\begin{equation}
N_{t+1} = S \times G + R.
\end{equation}

So, at each time step in a simulation, we use the survival regression to determine if each genet lives or not (if each quadrat remains occupied or not), the growth regression to determine size changes of surviving individuals (cover change of occupied quadrats), and the recruitment regression to determine the number of new recruits (if a quadrat is colonized or not). We first use one-step-ahead forecasts to assess each model's ability to reproduce observed cover changes. Then we use the models to analyze the effect of potential climate changes on population size over long time scales.

We used random draws from the final 1,000 iterations from each of 3 MCMC chains to introduce stochasticity into our population models. At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression. Then, we drew the full parameter set (specific random year effects, climate effects, fixed effects) from a randomly selected MCMC iteration. We selected random year effects and the climate year independently.

Results
-------
### Climate effects on vital rates
At the individual level we fit vital rate models that included climate covariates and random year effects (full model), just climate covariates (climate model), and no climate or random year effects (constant model). We used the metric from @Adler2012 to quantify the contribution of climate covariates to explaining interannual variability in vital rates based on comparisons of model deviance: (Climate model - Constant Model)/(Full model - Constant model). At the quadrat level, we could only fit random year effects for the growth regression. This is, in fact, one of the limitation of census-based data: local extirpation and colonization events are likely to be rare, making it difficult to fit complex statistical models. So, in this section we focus on growth only.

Climate covariates improved the growth regressions at both levels, with a slghty greater contribution at the individual level (Figure 1). Parameter estimates were generally consistent in terms of sign and magnitude between the individual level and the quadrat level for each species and climate covariate (Figure 2). There is no clear pattern suggesting that parameters estimated using individual level data are more certain, as evidenced by credible interval widths, than estimates based on quadrat data.

### Climate effects on population dynamics

Discussion
----------
We sought to test the assumption that the sensitivities of plant populations to climate variables can be detected equally well using either individual level data or population level data. This is an important question to answer because population models are key tools for predicting the consequences of global climate change. However, they can be of limited use when built on data from a small subset of a population in space or time. If population level data (i.e., some aggregated form of individual level data) can be used to detect climate effects on population dynamics, then we would have a cheaper and easier option for data collection over relatively large temporal and spatial extents [_e.g._ @Freckleton2011].

1. Climate detections
2. Limitations:
  * Sample size at quadrat level
  * are these species inherently sensitive to climate (compare to Chu and Adler in review -- say why we chose Montana site -- most sensitive to climate variables)
  * we know recruitment is important, and this is something that may be easily missed at quad level
3. Implications

Figures
-------

@@

gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
gIC <- gIC[133:148,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
gIC$X <- rep(coefNames, each=4)
gIC$rate <- "growth"
sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
sIC <- sIC[133:148,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
sIC$X <- rep(coefNames, each=4)
sIC$rate <- "survival"
rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
rIC <- rIC[81:96,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
rIC$X <- rep(coefNames, each=4)
rIC$rate <- "colonization"
plotIC <- rbind(gIC,sIC,rIC)
gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
  geom_hline(aes(yintercept=0), linetype="dashed")+
  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
  ylab("Standardized Coefficient Value")+
  xlab("Species")+
  facet_wrap(X~.)+
  coord_flip()+
  theme_bw()+
  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
                     labels=c("recruitment", "growth", "survival"))+
#   guides(color=FALSE)+
  guides(color=guide_legend(title=""))+
  theme(legend.position="top")+
  ggtitle("Individual level")

@@

```{r figure_4, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature. We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=4.5, fig.height=4, cache=FALSE, results='hide'}
climBase <- readRDS("../montana/quadBM/simulations/climateChangeBase.rds")
climPpt <- readRDS("../montana/quadBM/simulations/climateChangePrecip.rds")
climTemp <- readRDS("../montana/quadBM/simulations/climateChangeTemperature.rds")
climBase$sim <- rep("Observed", nrow(climBase)) 
climPpt$sim <- rep("Precipitation", nrow(climBase)) 
climTemp$sim <- rep("Temperature", nrow(climBase)) 
climSimsQBM <- rbind(climBase, climPpt, climTemp)
# ggplot(climSimsQBM)+
#   geom_boxplot(aes(x=species, y=cover*100, fill=sim))+
#   scale_y_continuous(limits=c(0,10))
climSimsQBMavg <- ddply(climSimsQBM, .(sim, species), summarise,
                        avgCover = median(cover))
climSimQBMcast <- dcast(climSimsQBMavg, formula = species~sim)
climSimQBMcast$diffP <- with(climSimQBMcast, (Precipitation*100)-(Observed*100)) 
climSimQBMcast$diffT <- with(climSimQBMcast, (Temperature*100)-(Observed*100))
climSimQBMplot <- melt(climSimQBMcast)
ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"))+
  geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge")+
  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase"), values=c("grey25", "grey50"))+
  theme_bw()+
  theme(legend.position=c(0.6, 0.8))+
  ylab("Cover Change (%)")+
  xlab("Species")
```

### 2014-12-29 manuscript/manuscript.tex
added climate tracking script
Added a climate tracking script, a la Jim Clark. Probably won’t use it
for this project, but could be interesting to look at all species at
all sites and examine climate tracking…
updated priors
new analysis update

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is
freely available on Ecological Archives (Anderson et al. 2011), and
interested readers should refer to the metadata therin for a complete
description. The site is about 800 m above sea level and mean annual
precipitation (1878-2009) is 334 mm, with most annual precipitaion
falling from April through September (76). The site is grass dominated
and, for the purposes of our study, we focus on the four most abundant
graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annualy
in 44 1-m2 quadrats using a pantograph. The quadrats were distributed in
six pastures, each assigned a grazing treatment of light (1.24 ha/animal
unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking
rates (two pastures per treatment). In this analysis we account for
potential differences among the grazing treatments, but do not focus on
grazing$\times$climate interactions. The annual maps of the quadrats
were digitized and the fates of individual plants tracked and extracted
using a computer program. Daily climate data, which we aggregated into
climate variables of interest, are available for the duration of the
data collection period (1932 - 1945) from the Miles City airport, Wiley
Field, 9 km from the study site.

@@

With the posterior distribution of the vital rate statistical models in
hand, it is straightforward to simulate the population models.

@@

@@

>>>>

\emph{Keywords}: forecasting, climate change, grassland, integral
projection model

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is
freely available on Ecological Archives (Anderson et al. 2011), and
interested readers should refer to the metadata therin for a complete
description. The site is about 800 m above sea level and mean annual
precipitation (1878-2009) is 334 mm, with most annual precipitaion
falling from April through September. The site is grass dominated and,
for the purposes of our study, we focus on the four most abundant
graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annualy
in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing$\times$climate interactions. The annual maps of the
quadrats were digitized and the fates of individual plants tracked and
extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

@@

With the posterior distribution of the vital rate statistical models in
hand, it is straightforward to simulate the population models. We used
an Integral Projection Model (IPM) to model populations based on
individual level data and an quadrat based version of an
individually-based model (Quadrat-Based Model, QBM) to model populations
based on quadrat level data. Both models take the general form:

\begin{equation}
N_{t+1} = S \times G + R.
\end{equation}

So, at each time step in a simulation, we use the survival regression to
determine if each genet lives or not (if each quadrat remains occupied
or not), the growth regression to determine size changes of surviving
individuals (cover change of occupied quadrats), and the recruitment
regression to determine the number of new recruits (if a quadrat is
colonized or not). We first use one-step-ahead forecasts to assess each
model's ability to reproduce observed cover changes. Then we use the
models to analyze the effect of potential climate changes on population
size over long time scales.

We used random draws from the final 1,000 iterations from each of 3 MCMC
chains to introduce stochasticity into our population models. At each
time step, we first randomly selected climate covariates from one of the
14 observed years and also randomly drew one set of random year effects
for each vital rate regression. Then, we drew the full parameter set
(specific random year effects, climate effects, fixed effects) from a
randomly selected MCMC iteration. We selected random year effects and
the climate year independently.

@@

\subsubsection{Climate effects on vital
rates}\label{climate-effects-on-vital-rates}

At the individual level we fit vital rate models that included climate
covariates and random year effects (full model), just climate covariates
(climate model), and no climate or random year effects (constant model).
We used the metric from Adler et al. (2012) to quantify the contribution
of climate covariates to explaining interannual variability in vital
rates based on comparisons of model deviance: (Climate model - Constant
Model)/(Full model - Constant model). At the quadrat level, we could
only fit random year effects for the growth regression. This is, in
fact, one of the limitation of census-based data: local extirpation and
colonization events are likely to be rare, making it difficult to fit
complex statistical models. So, in this section we focus on growth only.

Climate covariates improved the growth regressions at both levels, with
a slghty greater contribution at the individual level (Figure 1).
Parameter estimates were generally consistent in terms of sign and
magnitude between the individual level and the quadrat level for each
species and climate covariate (Figure 2). There is no clear pattern
suggesting that parameters estimated using individual level data are
more certain, as evidenced by credible interval widths, than estimates
based on quadrat data.

\subsubsection{Climate effects on population
dynamics}\label{climate-effects-on-population-dynamics}

\subsection{Discussion}\label{discussion}

We sought to test the assumption that the sensitivities of plant
populations to climate variables can be detected equally well using
either individual level data or population level data. This is an
important question to answer because population models are key tools for
predicting the consequences of global climate change. However, they can
be of limited use when built on data from a small subset of a population
in space or time. If population level data (i.e., some aggregated form
of individual level data) can be used to detect climate effects on
population dynamics, then we would have a cheaper and easier option for
data collection over relatively large temporal and spatial extents
(\emph{e.g.} Freckleton et al. 2011).

\subsection{Figures}\label{figures}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4.pdf}
\caption{Simulated changes species' median cover caused by a 1\%
increase in observed precipitation or temperature. We used the median of
simulated results because the average was highly influence by some
extreme outliers.}
\end{figure}
